---
title: "Using rSEA to analyze pathway enrichment for CPTAC data"
author: "Caleb Lindgren"
---

# Using rSEA to analyze pathway enrichment for CPTAC data

## Setup
```{r, message=FALSE}
library(tidyverse)
library(rSEA)

TIME_START = format(Sys.time(), "%Y%m%d_%H%M%S")

STEP01_DIR <- "step01_outputs"
STEP01_FILE <- "tumor_change_20200529_195104_10000_perm.tsv"
STEP01_FILE_PATH <- file.path(STEP01_DIR, STEP01_FILE)

PATHWAYS_DIR <- "gene_set_libraries"
PATHWAYS_FILE <- "Reactome_2020.gmt"
PATHWAYS_FILE_PATH <- file.path(PATHWAYS_DIR, PATHWAYS_FILE)

STEP16_DIR <- "step16_outputs"
STEP16_FILE <- paste("enrichment_rsea_", TIME_START, ".tsv", sep = "")
STEP16_FILE_PATH <- file.path(STEP16_DIR, STEP16_FILE)

# Create the directory if it doesn't already exist
dir.create(STEP16_DIR, showWarnings = FALSE)
```

## Step 1: Load expression data and divide by cancer type

We will use a list of genes, with permutation-based *p*-values for differential expression between tumor and normal, that we generated previously.

```{r, message=FALSE, warning=FALSE}
all_expr_data <- read_tsv(STEP01_FILE_PATH)
cancer_types <- distinct(all_expr_data, cancer_type)$cancer_type
split_expr_data <- lapply(cancer_types, function(x) filter(all_expr_data, cancer_type == x))
names(split_expr_data) <- cancer_types
```

## Step 2: Load pathway lists

We have the Reactome pathways stored as a .gmt file. We will read this in as a list of lists.

```{r}
pathways <- PATHWAYS_FILE_PATH %>%
  readLines() %>% # Read in each line of the file as a string
  sapply(strsplit, split = "\t") %>% # Split each line on the tab character
  sapply(function(x) x[-1]) # Cut out the pathway names

pathway_ids <- sapply(pathways, function(x) x[[1]]) # Save the pathway ids
pathways <- sapply(pathways, function(x) x[-1]) # Cut out the pathway ids from the lists
names(pathways) <- pathway_ids # Set the pathway ids as the names
```

## Step 3: Run rSEA and save the results
```{r}
results <- lapply(split_expr_data, function(x) SEA(x$P_val, x$protein_str, pathlist = pathways))
names(results) <- cancer_types

results <- cancer_types %>%
  lapply(function(x) results[[x]] %>% add_column(cancer_type = x)) %>% # Add each table's cancer type as a column
  bind_rows() # Combine all the tables into one table

write_tsv(results, STEP16_FILE_PATH)
```